Tools for identifying unexpectedly low microfossil count sums

### 2020-06-11 Rmd/count_check_MS.Rmd
pollen2 pollen3 text

Dataset pollen1, published includes `r pollenA$n_samp` assemblages with a median richness of `r pollenA$taxa_median` taxa.
Count sums vary between `r pollenA$count_min` and `r pollenA$count_max`.
The greatest common divisor is three for all assemblages: all `r pollenA$n_count` counts in the dataset are divisible by three. 
Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$.
For a dataset this large, the probability of having all counts divisible by three by chance is 10^`r ceiling(- pollenA$n_count * log10(3))`^.
Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts, but I have no idea how they were generated.

All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.
The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assemblages with the expected greatest common divisor of one.
The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table)).

@@

Assemblages with a greatest common divisor of one typically have a higher taxonomic richness than assemblages with a greatest common divisor of two (Fig. \@ref(fig:neotoma-plot)).
In the datasets with a wide range of count sums, the richness of the assemblages with a greatest common divisor of two is comparable with the richness of the assemblages with a greatest common divisor of one that have a count sum that is half as large.

```{r neotoma-plot, fig.height = 70/24.5, fig.width=80/24.5, fig.cap="Number of taxa in against count sum for assemblages in two pollen datasets with a greatest common divisor of one (red circles) or two (blue triangles). Lines are loess with a span of 1. All four pollen datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two.", message = FALSE}
pollen1234$figure %>% print()
```

>>>>

Dataset pollen1, published includes `r pollenA$n_samp` assemblages with a median richness of `r pollenA$taxa_median` taxa.
Count sums vary between `r pollenA$count_min` and `r pollenA$count_max`.
The greatest common divisor is three for all assemblages: all `r pollenA$n_count` counts in the dataset are divisible by three. 
Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $k^{-n}$.
For a dataset this large, the probability of having all counts divisible by three by chance is 10^`r ceiling(- pollenA$n_count * log10(3))`^.
Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts.

In datasets pollen2 and pollen3, presented in the same paper, # % and # % of assemblages do not have singletons and have a greatest common divisor greater than one.
Both datasets match the published pollen stratigraphies.
For most of both datasets, assemblages with and without singletons are arranged alternately.
Samples without singletons are on the 10 cm intervals, and those with on the intervening 5 cm intervals. 
Closer inspection reveals that the assemblages between two assemblages without singletons have counts that are the mean of their neighbours.
On request, the authors sent the original count data. 
Comparison of these with the archived data showed that the archived assemblages without singletons had had their counts multiplied by two.
The intervening assemblages did not exist in the original data. 
In the published and archived data, they had been interpolated from their neighbours (Fig. \@ref(fig:archived-original-plot-plot)).

fig - plot of 1 or 2 mid-common taxa by depth in archived and original

@@

```{r archived-original-plot, fig.height = 70/24.5, fig.width=80/24.5, fig.cap="Number of taxa in against count sum for assemblages in two pollen datasets with a greatest common divisor of one (red circles) or two (blue triangles). Lines are loess with a span of 1. All four pollen datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two.", message = FALSE}
readd(archived_original_plot)
```
